---
title: "Data Exploration and Cleaning"
author: "Sophia Bessias"
date: "January 31, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Description
Script 1/4. This script cleans the dataset and splits it into training, test, and validation. 

####Load packages
```{r}
require(tidyverse)
require(caTools)
require(caret)
```

###Set path, filename
```{r}
path <- 'C:/Users/sbessias/Google Drive/NCSU/Applications/RTI/Project/'
filename <- 'master.csv'
```

###Load data
```{r, echo=F}
m <- read_csv(file=paste(path, filename, sep=''), na=c("", "?", "NA"))
read_csv
```

###Check for missing values
3 variables have >1% missing values: workclass, occupation, country. No variable has >6% missing values.
```{r}
vars <- colnames(m)
n = length(vars)
missing_prop = tibble(var_name = vars, p_miss = rep(0,n))

for (i in 1:n){
  missing_prop[i,2] = (sum(is.na(m[,i]))/nrow(m))
}

#see variables missing >1%
missing_prop[missing_prop$p_miss > 0.01 , ]
```

###Explore continuous variables
Check distributions, missing values
```{r}
#check for duplicate id values
#number of unique id's is the same as the number of rows in m:
length(unique(m$id))
dim(m)

#see summary of each continous variable
lapply(m[2:5], summary)

#histogram of capital gain where capital_gain > 0
hist(m$capital_gain[m$capital_gain > 0])

#histogram of capital gain  where capital_loss > 0
hist(m$capital_loss[m$capital_loss > 0])

```

###Explore factor variables
```{r}
#view tables for factor variables only (columns 6:14)
lapply(m[6:14], table)

```

### Data Cleaning
```{r}
#problem: low frequencies observed for workclass: never-worked (10), without-pay (21)
#solution: treat as missing
m$workclass_clean <- ifelse(m$workclass %in% c('Never-worked', 'Without-pay'), NA, m$workclass)

#problem: low frequencies observed for marital_status: Married-AF-Spouse (37)
#solution: treat as missing
m$marital_status_clean <- ifelse(m$marital_status == 'Married-AF-spouse', NA, m$marital_status)

#problem: low frequencies observed for occupation: Armed-forces(15)
#solution: treat as missing
m$occupation_clean <- ifelse(m$occuption == 'Armed-Forces', NA, m$occuption)

#problem: too many levels for highest_edu
#solution: combine into standard ranges
m$highest_edu_clean <- recode(m$highest_edu, 
                 'Preschool' = 'Elementary-or-Pre',
                 '1st-4th' = 'Elementary-or-Pre',
                 '5th-6th' = 'Middle-school',
                 '7th-8th' = 'Middle-school',
                 '9th' = 'Some-HS',
                 '10th' = 'Some-HS',
                 '11th' = 'Some-HS',
                 '12th' = 'Some-HS',
                 'Assoc-acdm' = 'Assoc',
                 'Assoc-voc' = 'Assoc',
                 'Masters' = 'Graduate-school',
                 'Doctorate' = 'Graduate-school'
                 )

#problem: too many levels for country
#solution: combine into UN categories
#https://en.wikipedia.org/wiki/United_Nations_Regional_Groups

m$world_region <- recode(m$country,
                          'Cambodia' = 'Asia-Pacific',
                          'Canada' = 'Western-European-and-Others',
                          'China' = 'Asia-Pacific',
                          'Columbia' = 'Latin-American-and-Caribbean',
                          'Cuba' = 'Latin-American-and-Caribbean',
                          'Dominican-Republic' = 'Latin-American-and-Caribbean',
                          'Ecuador' = 'Latin-American-and-Caribbean',
                          'El-Salvador' = 'Latin-American-and-Caribbean',
                          'England' = 'Western-European-and-Others',
                          'France' = 'Western-European-and-Others',
                          'Germany' = 'Western-European-and-Others',
                          'Greece' = 'Western-European-and-Others',
                          'Guatemala' = 'Latin-American-and-Caribbean',
                          'Haiti' = 'Latin-American-and-Caribbean',
                          'Holand-Netherlands' = 'Western-European-and-Others',
                          'Honduras' = 'Latin-American-and-Caribbean',
                          'Hong' = 'Asia-Pacific',
                          'Hungary' = 'Eastern-Europe',
                          'India' = 'Asia-Pacific',
                          'Iran' = 'Asia-Pacific',
                          'Ireland' = 'Western-European-and-Others',
                          'Italy' = 'Western-European-and-Others',
                          'Jamaica' = 'Latin-American-and-Caribbean',
                          'Japan' = 'Asia-Pacific',
                          'Laos' = 'Asia-Pacific',
                          'Mexico' = 'Latin-American-and-Caribbean',
                          'Nicaragua' = 'Latin-American-and-Caribbean',
                          'Outlying-US(Guam-USVI-etc)' = 'US-territories',
                          'Peru' = 'Latin-American-and-Caribbean',
                          'Philippines' = 'Asia-Pacific',
                          'Poland' = 'Eastern-Europe',
                          'Portugal' = 'Western-European-and-Others',
                          'Puerto-Rico' = 'US-territories',
                          'Scotland' = 'Western-European-and-Others',
                          'South' = 'Asia-Pacific',
                          'Taiwan' = 'Asia-Pacific',
                          'Thailand' ='Asia-Pacific' ,
                          'Trinadad&Tobago' = 'Latin-American-and-Caribbean',
                          'Vietnam' = 'Asia-Pacific',
                          'Yugoslavia' = 'Eastern-Europe')

```

###Split into training, test, and validation sets
```{r}
#set seed for reproducibility
set.seed(919)

#create index
idx <- sample(seq(1, 3), size = nrow(m), replace = T, prob = c(.7, .15, .15))

#split
train <- m[idx == 1,]
test <- m[idx == 2,]
val <- m[idx == 3,]

#save files
write_csv(m, path=paste(path, 'all.csv', sep=''))
write_csv(train, path=paste(path, 'train.csv', sep=''))
write_csv(test, path=paste(path, 'test.csv', sep=''))
write_csv(val, path=paste(path, 'val.csv', sep=''))
```

###Explore two-way associations using training data
All factor vars appear significantly associated with the target variable at alpha = 0.05
```{r}
factor_set =train[, c(6, 12:13, 15:19)]
vars = colnames(factor_set)
n = length(vars)
chisq_p = tibble(var_name = vars, p_val = rep(0,n))

for (i in 1:n){
  chitest = chisq.test(factor_set[,i], factor_set$over_50k)
  chisq_p[i,2] = chitest$p.value
}

chisq_p

```